---
title: "UpstreamDataCleaning"
output: html_document
---
```{r}
library(stringr)
library(tidyverse)
```

```{r}
rawSpecimenGroup <- read.csv("../BeatAML_reports/specimen_group_dump_20170306.csv")
rawLabResults <- read.csv("../BeatAML_reports/lab_results_dump_20170306.csv", na.strings = c("Not Available", "Not available", "not available", "N/A"))
rawDiagnosis <- read.csv("../BeatAML_reports/diagnosis_dump_20170306.csv")
```

Filter down to only Initial Acute Leukemia Diagnoses AND BeatAML Datalock

```{r}
initialDiagnosisSample <- rawSpecimenGroup %>% filter(Specimen.Groups..Specimen.Group == "Initial Acute Leukemia Diagnosis")
datalock1AML <- rawSpecimenGroup %>% 
  filter(Specimen.Groups..Specimen.Group == "DataLock1_TumorSample_NoMatchingNormal" |
           Specimen.Groups..Specimen.Group == "DataLock1_TumorSample_WithMatchingNormal")

dl1AMLlist <- initialDiagnosisSample[initialDiagnosisSample$Specimen..Lab.ID %in% datalock1AML$Specimen..Lab.ID,]
dl1AMLlist <- dl1AMLlist %>% filter(Specimen..Lab.ID != "15-00821") # Removes a bad sample, listed as both initial diagnosis and post-chemo, conflicting karyotypes
#nrow(dl1AMLlist)
```

Filter to lab results of only selected samples
```{r}
selectedLabs <- rawLabResults[rawLabResults$Specimen..Lab.ID %in% dl1AMLlist$Specimen..Lab.ID,]
```

Extract Relevant measures
```{r}
## Karyotype

selected_k <- selectedLabs %>%
  filter(Labs..Lab.Type == "Karyotype") %>% 
  select(Specimen..Lab.ID, Labs..Result) %>% 
  rename(CYTOGENETICS = Labs..Result)
  

dl1AMLlist_k <- left_join(dl1AMLlist, selected_k, by = "Specimen..Lab.ID")


## FLT3-ITD
flt.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "FLT3-ITD")) %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

flt.results$Specimen..Lab.ID <- factor(flt.results$Specimen..Lab.ID)
  
flt.wide <- flt.results %>% 
    unique() %>% 
    spread(Labs..Lab.Type, Labs..Result) %>% 
    mutate_each(funs(as.character(.)), 2:4)


flt.clean <- flt.wide %>% 
  mutate(priorityFLT3 = 
       ifelse(!is.na(flt.wide$`FLT3-ITD (internal)`), flt.wide$`FLT3-ITD (internal)`,
              ifelse(!is.na(flt.wide$`FLT3-ITD (GeneTrails)`), flt.wide$`FLT3-ITD (GeneTrails)`,
                     ifelse(!is.na(flt.wide$`FLT3-ITD`), flt.wide$`FLT3-ITD`, NA)))
       ) %>% 
  mutate(FLT3_ITD = 
           ifelse(priorityFLT3 == "Positive", TRUE, 
                  ifelse(priorityFLT3 == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, FLT3_ITD)

dl1AMLlist_kf <- left_join(dl1AMLlist_k, flt.clean, by = "Specimen..Lab.ID")


## NPM1

npm.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "NPM")) %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

npm.results$Specimen..Lab.ID <- factor(npm.results$Specimen..Lab.ID)
  
npm.wide <- npm.results %>% 
    unique() %>% 
    spread(Labs..Lab.Type, Labs..Result) %>% 
    mutate_each(funs(as.character(.)), 2:4)

npm.clean <- npm.wide %>% 
  mutate(priorityNPM1 = 
       ifelse(!is.na(npm.wide$`NPM1 (internal)`), npm.wide$`NPM1 (internal)`,
              ifelse(!is.na(npm.wide$`NPM (GeneTrails)`), npm.wide$`NPM (GeneTrails)`,
                     ifelse(!is.na(npm.wide$NPM), npm.wide$NPM, NA)))
       ) %>% 
  mutate(NPM1_MUTATED = 
           ifelse(priorityNPM1 == "Positive", TRUE, 
                  ifelse(priorityNPM1 == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, NPM1_MUTATED)

dl1AMLlist_kfn <- left_join(dl1AMLlist_kf, npm.clean, by = "Specimen..Lab.ID")

## CEBPA

cebpa.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "CEBPA")) %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

cebpa.results$Specimen..Lab.ID <- factor(cebpa.results$Specimen..Lab.ID)
  
cebpa.wide <- cebpa.results %>% 
    unique() %>% 
    spread(Labs..Lab.Type, Labs..Result) %>% 
    mutate_each(funs(as.character(.)), 2:3)

cebpa.clean <- cebpa.wide %>% 
  mutate(priorityCEBPA = 
       ifelse(!is.na(cebpa.wide$`CEBPA (GeneTrails)`), cebpa.wide$`CEBPA (GeneTrails)`,
              ifelse(!is.na(cebpa.wide$CEBPA), cebpa.wide$CEBPA, NA))
       ) %>% 
  mutate(CEBPA_MUTATED = 
           ifelse(priorityCEBPA == "Positive", TRUE, 
                  ifelse(priorityCEBPA == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, CEBPA_MUTATED)

dl1AMLlist_kfnc <- left_join(dl1AMLlist_kfn, cebpa.clean, by = "Specimen..Lab.ID")

## TP53

TP53.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "TP53")) %>%
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

TP53.results$Specimen..Lab.ID <- factor(TP53.results$Specimen..Lab.ID)

TP53.wide <- TP53.results %>% 
    unique() %>% 
    spread(Labs..Lab.Type, Labs..Result) %>% 
    mutate_each(funs(as.character(.)), 2:3)

TP53.clean <- TP53.wide %>% 
  mutate(priorityTP53 = 
       ifelse(!is.na(TP53.wide$`TP53 (GeneTrails)`), TP53.wide$`TP53 (GeneTrails)`,
              ifelse(!is.na(TP53.wide$TP53), TP53.wide$TP53, NA))
       ) %>% 
  mutate(TP53 = 
           ifelse(priorityTP53 == "Positive", TRUE, 
                  ifelse(priorityTP53 == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, TP53)

## RUNX1

RUNX1.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "RUNX1")) %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

RUNX1.results$Specimen..Lab.ID <- factor(RUNX1.results$Specimen..Lab.ID)

RUNX1.wide <- RUNX1.results %>% 
  unique() %>% 
  spread(Labs..Lab.Type, Labs..Result) %>% 
  mutate_each(funs(as.character(.)), 2:3)

RUNX1.clean <- RUNX1.wide %>% 
  mutate(priorityRUNX1 =
           ifelse(!is.na(RUNX1.wide$`RUNX1 (GeneTrails)`), RUNX1.wide$`RUNX1 (GeneTrails)`,
              ifelse(!is.na(RUNX1.wide$RUNX1), RUNX1.wide$RUNX1, NA))
       ) %>% 
  mutate(RUNX1 = 
           ifelse(priorityRUNX1 == "Positive", TRUE, 
                  ifelse(priorityRUNX1 == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, RUNX1)

## ASXL1

ASXL1.results <- selectedLabs %>% 
  filter(str_detect(Labs..Lab.Type, "ASXL1")) %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type) %>% 
  arrange(Specimen..Lab.ID)

ASXL1.results$Specimen..Lab.ID <- factor(ASXL1.results$Specimen..Lab.ID)

ASXL1.wide <- ASXL1.results %>% 
  unique() %>% 
  spread(Labs..Lab.Type, Labs..Result) %>% 
  mutate_each(funs(as.character(.)), 2:3)

ASXL1.clean <- ASXL1.wide %>% 
  mutate(priorityASXL1 =
           ifelse(!is.na(ASXL1.wide$`ASXL1 (GeneTrails)`), ASXL1.wide$`ASXL1 (GeneTrails)`,
              ifelse(!is.na(ASXL1.wide$ASXL1), ASXL1.wide$ASXL1, NA))
       ) %>% 
  mutate(ASXL1 = 
           ifelse(priorityASXL1 == "Positive", TRUE, 
                  ifelse(priorityASXL1 == "Negative", FALSE, NA))) %>% 
  select(Specimen..Lab.ID, ASXL1)

dl1AMLlist_kfncpra <- left_join(dl1AMLlist_kfnc, TP53.clean, by = "Specimen..Lab.ID") %>% 
  left_join(., RUNX1.clean, by = "Specimen..Lab.ID") %>% 
  left_join(., ASXL1.clean, by = "Specimen..Lab.ID")
```

Find Genetrails = negative fields, this means genetrails was run and no mutations found, so it should trump NA fields with "Negative"

```{r}
#Way too many GT = positive values... going to have to make sure they are properly scored
#Negative values first
genetrails <- selectedLabs %>% 
  filter(Labs..Lab.Type == "GeneTrails") %>% 
  select(Specimen..Lab.ID, Labs..Result, Labs..Lab.Type, Labs..Comment) %>% 
  arrange(Specimen..Lab.ID)

genetrails$Specimen..Lab.ID <- factor(genetrails$Specimen..Lab.ID)

genetrails.wide <- genetrails %>% 
  spread(Labs..Lab.Type, Labs..Result) %>% 
  mutate_each(funs(as.character(.)), 3)
  
genetrails.clean <- genetrails.wide %>% 
  mutate(GENETRAILS_PANEL_GENERAL = 
           ifelse(GeneTrails == "positive", "Positive",genetrails.wide$GeneTrails)) %>% 
  select(Specimen..Lab.ID, GENETRAILS_PANEL_GENERAL, Labs..Comment)

genetrails.clean$GENETRAILS_PANEL_GENERAL <- factor(genetrails.clean$GENETRAILS_PANEL_GENERAL)

summary(genetrails.clean)

dl1AMLlist_kfncprag <- left_join(dl1AMLlist_kfncpra, genetrails.clean, by = "Specimen..Lab.ID")
compiled_data <- dl1AMLlist_kfncprag
```

## Fixing mismatched data
### If Genetrails != NA, AND FLT/NPM/CEBPA = NA, change to negative
I manually confirmed that none of the genetrails comments report a mutation that is listed as NA in the gene-specific field, including P53, RUNX1, ASXL1
```{r}
nrow(discrepencies)

compiled_data_neg <- compiled_data %>%
  mutate(FLT3_ITD = 
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL) 
                  & is.na(compiled_data$FLT3_ITD), FALSE, compiled_data$FLT3_ITD)
         ) %>% 
  mutate(NPM1_MUTATED = 
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL) 
                  & is.na(compiled_data$NPM1_MUTATED), FALSE, compiled_data$NPM1_MUTATED)
           ) %>% 
  mutate(CEBPA_MUTATED = 
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL) 
                  & is.na(compiled_data$CEBPA_MUTATED), FALSE, compiled_data$CEBPA_MUTATED)
           ) %>% 
  mutate(TP53 =
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL)
                  & is.na(compiled_data$TP53), FALSE, compiled_data$TP53)
         ) %>% 
  mutate(RUNX1 =
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL)
                  & is.na(compiled_data$RUNX1), FALSE, compiled_data$RUNX1)
         ) %>%  
  mutate(ASXL1 =
           ifelse(!is.na(compiled_data$GENETRAILS_PANEL_GENERAL)
                  & is.na(compiled_data$ASXL1), FALSE, compiled_data$ASXL1)
         )


compiled_data_neg %>%
  filter(!is.na(GENETRAILS_PANEL_GENERAL)) %>% 
  filter(is.na(FLT3_ITD) | is.na(NPM1_MUTATED) | is.na(CEBPA_MUTATED) |
           is.na(TP53) | is.na(RUNX1) | is.na(ASXL1)) %>% 
  select(Specimen..Lab.ID,FLT3_ITD, NPM1_MUTATED, CEBPA_MUTATED,
         TP53, RUNX1, ASXL1, GENETRAILS_PANEL_GENERAL, Labs..Comment) -> discrepencies

nrow(discrepencies)

all_data <- compiled_data_neg

complete_records <- all_data %>% 
  filter(!is.na(CYTOGENETICS) &
                      !is.na(FLT3_ITD) &
                      !is.na(NPM1_MUTATED) &
                      !is.na(CEBPA_MUTATED) &
                      !is.na(RUNX1) &
                      !is.na(ASXL1) &
                      !is.na(TP53)) %>% 
  rename(lab_id = Specimen..Lab.ID) %>% 
  rename(patient_id = Patient..Patient.ID) %>% 
  select(-Specimen.Groups..Specimen.Group, -GENETRAILS_PANEL_GENERAL, -Labs..Comment) %>% 
  rename(NPM1 = NPM1_MUTATED) %>% 
  rename(CEBPA = CEBPA_MUTATED)

complete_records %>% nrow()
complete_records %>% head()

write.csv(complete_records, "./data/BeatAMLdataset.csv")
```




